Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-06-03 paper/paper.qmd
Reduce abstract < 250 words

---
title: "Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome"
date: last-modified
date-format: medium
prefer-html: true
bibliography: references.bib
fontsize: 11pt
format:
  manuscript-pdf:
    keep-tex: true
    keep-md: true
    linenumbers: true
    doublespacing: true
    lof: false
  html:
    embed-resources: true
runningtitle: Predicting CDI Severity from OTUs
runningauthor: Sovacool
papertype: Preprint
author:
  - name: Kelly L. Sovacool
    orcid: 0000-0003-3283-829X
    affiliations:
      - ref: bioinf
  - name: Sarah E. Tomkovich
    affiliations:
      - ref: micro
  - name: Megan L. Coden
    affiliations:
      - ref: mcdb
  - name: Vincent B. Young
    affiliations:
      - ref: micro
      - ref: intmed
  - name: Krishna Rao
    affiliations:
      - ref: intmed
  - name: Patrick D. Schloss
    orcid: 0000-0002-6935-4275
    email: pschloss@umich.edu
    corresponding: true
    affiliations:
      - ref: micro
      - ref: ccmb
affiliations:
  - id: bioinf
    name: Department of Computational Medicine & Bioinformatics, University of Michigan
  - id: micro
    name: Department of Microbiology & Immunology, University of Michigan
  - id: mcdb
    name: Department of Molecular, Cellular, and Developmental Biology, University of Michigan
  - id: intmed
    name: Division of Infectious Diseases, Department of Internal Medicine, University of Michigan
  - id: ccmb
    name: Center for Computational Medicine and Bioinformatics, University of Michigan
abstract: |
  _Clostridioides difficile_ infection (CDI) can lead to 
  adverse outcomes including ICU admission, colectomy, and death.
  The composition of the gut microbiome plays an important role in determining
  colonization resistance and clearance upon exposure to _C. difficile_.
  We investigated whether machine learning (ML) models trained on gut microbiota
  isolated from stool samples on the day of CDI diagnosis could predict which
  CDI cases led to severe outcomes.
  We collected 1,277 stool samples from CDI patients,
  performed 16S rRNA gene amplicon sequencing, then processed and clustered the
  sequences into _de novo_ Operational Taxonomic Units.
  We trained ML models to predict CDI severity according to four different
  severity definitions: 
  the IDSA severity score on the day of diagnosis,
  all-cause adverse outcomes within 30 days,
  disease-related complications confirmed attributable to CDI via chart review, 
  and a pragmatic definition that uses the attributable definition when available
  and falls back to the all-cause definition otherwise.
  The models predicting the pragmatic severity definition performed best, 
  suggesting that while chart review is valuable to verify the cause of
  complications, including as many samples as possible is indispensable for
  training performant models on imbalanced datasets.
  Permutation importance identified _Enterococcus_ as the most important OTU for
  model performance across all severity definitions. 
  Finally, we evaluated the potential clinical value of the OTU-based models
  and found similar performance compared to prior models based on Electronic
  Health Record data.
  Our results show that OTU-based models can modestly identify patients at risk
  for severe CDI, which could be used by healthcare providers to identify
  patients at risk and ultimately improve CDI outcomes.
#importance: TODO
keywords: _C. difficile_ infection, supervised machine learning, gut microbiome, amplicion sequencing
execute:
  eval: true
  echo: false
  output: false
  cache: true
  message: false
  warning: false
---
```{r setup}
schtools::set_knitr_opts()
knitr::opts_chunk$set(dpi = 600)
```
```{r load_data}
library(assertthat)
library(glue)
library(here)
library(kableExtra)
library(knitr)
library(tidyverse)

>>>>

---
title: "Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome"
date: last-modified
date-format: medium
prefer-html: true
bibliography: references.bib
fontsize: 11pt
format:
  manuscript-pdf:
    keep-tex: true
    keep-md: true
    linenumbers: true
    doublespacing: true
    lof: false
  html:
    embed-resources: true
runningtitle: Predicting CDI Severity from OTUs
runningauthor: Sovacool
papertype: Preprint
author:
  - name: Kelly L. Sovacool
    orcid: 0000-0003-3283-829X
    affiliations:
      - ref: bioinf
  - name: Sarah E. Tomkovich
    affiliations:
      - ref: micro
  - name: Megan L. Coden
    affiliations:
      - ref: mcdb
  - name: Vincent B. Young
    affiliations:
      - ref: micro
      - ref: intmed
  - name: Krishna Rao
    affiliations:
      - ref: intmed
  - name: Patrick D. Schloss
    orcid: 0000-0002-6935-4275
    email: pschloss@umich.edu
    corresponding: true
    affiliations:
      - ref: micro
      - ref: ccmb
affiliations:
  - id: bioinf
    name: Department of Computational Medicine & Bioinformatics, University of Michigan
  - id: micro
    name: Department of Microbiology & Immunology, University of Michigan
  - id: mcdb
    name: Department of Molecular, Cellular, and Developmental Biology, University of Michigan
  - id: intmed
    name: Division of Infectious Diseases, Department of Internal Medicine, University of Michigan
  - id: ccmb
    name: Center for Computational Medicine and Bioinformatics, University of Michigan
abstract: |
  _Clostridioides difficile_ infection (CDI) can lead to 
  adverse outcomes including ICU admission, colectomy, and death.
  The composition of the gut microbiome plays an important role in determining
  colonization resistance and clearance upon exposure to _C. difficile_.
  We investigated whether machine learning (ML) models trained on gut microbiota
  isolated from stool samples on the day of CDI diagnosis could predict which
  CDI cases led to severe outcomes.
  We collected 1,277 stool samples from CDI patients and clustered 16S rRNA gene
  amplicon sequences into _de novo_ Operational Taxonomic Units (OTUs).
  We then trained ML models to predict CDI severity on OTU relative abundances
  according to four different severity definitions: 
  the IDSA severity score on the day of diagnosis,
  all-cause adverse outcomes within 30 days,
  adverse outcomes confirmed as attributable to CDI via chart review, 
  and a pragmatic definition that uses the attributable definition when available
  and falls back to the all-cause definition otherwise.
  The models predicting pragmatic severity performed best, 
  suggesting that while chart review is valuable to verify the cause of
  complications,
  including as many samples as possible is indispensable for training performant
  models on imbalanced datasets.
  Permutation importance identified _Enterococcus_ as the most important OTU for
  model performance. 
  Finally, we evaluated the potential clinical value of the OTU-based models
  and found similar performance compared to prior models based on Electronic
  Health Records.
  The modest performance of the OTU-based models represents a step toward the
  goal of deploying models to inform clinical decisions and ultimately improve
  CDI outcomes.
#importance: TODO
keywords: _C. difficile_ infection, supervised machine learning, gut microbiome, amplicion sequencing
execute:
  eval: true
  echo: false
  output: false
  cache: true
  message: false
  warning: false
---
```{r setup}
schtools::set_knitr_opts()
knitr::opts_chunk$set(dpi = 600)
```
```{r load_data}
library(assertthat)
library(glue)
library(here)
library(kableExtra)
library(knitr)
library(tidyverse)
